On Optimality of Finite Element Integration

### 2015-12-21 TOMS_Paper/assembly.tex
Rework, again, the algorithm...including the sharing graph

\subsubsection{Structured tensor operations}
\label{sec:se-rln}
Finite element expressions can be seen as composition of operations between tensors. We observe that, when multiplying two tensors, it often happens that the optimal scheduling strategy is to be searched in a space of size 2. To illustrate the concept, we take a matrix-vector product $y_{\Psi \Delta} = A_{\Psi} x_{\Delta}$, in which the entries of $y, A$ and $x$ depend on $\Psi, \Delta \subset \Lambda$. Let us assume that $y_{\Psi \Delta}$ itself is an operand of a larger scalar-valued expression, so its entries can represent any possible sub-expression in $\lbrace \alpha_{eij}, \beta_{eik}, \sigma_{ei} \rbrace$. For example, suppose that $A$ is the inverse Jacobian matrix of a given coefficient and $x$ the gradient of a function in two dimensions, and consider the case $\Psi = \emptyset$, $\Delta = \lbrace L_i \rbrace$. Thus an entry in $y_{L_i}$ is of the form $(a f_i + b g_i) (c f_i + d g_i) \in \sigma_{ei}$. To schedule this expression, we basically have two options:
\begin{enumerate}[label=(\roman*)]
\item Applying generalized code motion -- not needed in the example as we have one loop.
\item Searching for spatial sharing, expanding and factorizing the expression to promote temporal sharing, applying generalized code motion -- the expression in the example would be recast as $f_i (a + c) + g_i (b + d)$, with $(a + c)$ and $(b + d)$ being two hoistable sub-expressions.
\end{enumerate}
The reader can verify that (ii) improves the operation count over (i) for any size of $L_i$. In general, however, the optimal option depends on multiple factors: the loop size, the expansion cost, and the factorization gain. One can appreciate the dichotomy between (i) and (ii) in several expressions originating from a wide range of variational forms\footnote{In fact, it systematically appears in all of the forms used for experimentation (Section~\ref{sec:perf-results}), and many more.}. We then regard as \textit{structured tensor operation} any operation between two tensors that result in sub-expressions for which there is no ambiguity in spatial sharing elimination -- implying that the optimal operation scheduling is to be determined out of two alternatives.

@@

The algorithm makes use of the following notation and terminology:
\begin{itemize}
\item $|.|$: a generic ``sizeof'' operator (e.g., cardinality of a set, extent of a loop).
\item \textit{multilinear operand}: any $\alpha_{eij}$ or $\beta_{eik}$ in the input expression.
\item \textit{multilinear symbol}: a symbol appearing within a multilinear operand depending on $L_j$ or $L_k$ (e.g., test functions, first order derivatives of test functions, etc.).
\end{itemize}

\begin{Algo}[Sharing elimination]
\label{algo:sharing-elimination}
\normalfont
The algorithm has three main phases: initialization (step 1); scheduling of multilinear operands for reaching quasi-optimality (steps 2-5); scheduling of all other sub-expressions (step 6).
\begin{enumerate}
\item Perform a depth-first visit of the expression tree with recursive collection of identical sub-expressions.

\textit{Note: for example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$}.
\item Perform a depth-first visit of the expression tree to collect and partition multilinear operands into $p$ disjoint sets, $\mathbb{P} = \lbrace P^1, ..., P^{p}\rbrace$. $\mathbb{P}$ is such that multilinear operands in $P$ share a set of multilinear symbols $S_P$, whereas there is no sharing across different partitions.
\item Calculate the number of operations in the innermost loop required for the execution of $P \in \mathbb{P}$. There are two possibilities, depending on which strategy -- (i) or (ii), as described in Section~\ref{sec:se-rln} -- is used. In particular, we have $c_P = {min(}c_P^{i},\ c_P^{ii}{)}$, with $c_P^i = |P|$ and $c_P^{ii} = |S_P|$.
\item For all multilinear operands $o \in P$ such that $c_P = c_P^i$, apply strategy (i). Update $\mathbb{P}$ by replacing hoisted multilinear operands with the corresponding temporary symbols.
\item For all $P \in \mathbb{P}$, determine the profitability of strategy (ii). We introduce $g_P$, which measures the gain in operation count in the innermost loop if (ii) were applied. We have $g_P = \sum_{Q \in \mathbb{Q}} c_{Q_{old}} - c_{Q_{new}}$, with $\mathbb{Q} = \mathbb{P} \setminus \lbrace P \rbrace$. Apply (ii) if $g_P > 0$ and the expansion cost does not exceed the gain in operation count.

\textit{Note: both $c_{Q_{new}}$ and the expansion cost are easily determined through tree visits.}

\textit{Note: $\mathbb{P}$ is in practice very small, so this pass can be accomplished rapidly.}
\item Perform a depth-first visit of the expression tree and, for each $L_i$-dependent or previously hoisted sub-expression, apply the most profitable between (i) and (ii).

\textit{Note: this pass speculatively assumes that structured tensor operations are in place. If the assumption does not hold, the result will generally be sub-optimal since only a subset of code motion opportunities may have been exposed.}
\end{enumerate}
\end{Algo}

%TODO : dire che il costo dell'espansione è semplicemente dato da ciascun s in | S_P | times la 2 evalato al depth di quell s.

\subsubsection{Examples}
\label{sec:se-examples}
Figure~\ref{code:poisson} applies Algorithm~\ref{algo:sharing-elimination} to the bilinear form arising from a Poisson equation in 2D. We observe that $P_k^0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P_j^1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$, with $c_{P^0} = c_{P^1} = {min(}2, 2{)} = 2$. A much more complex example, using the hyperelastic model evaluated in Section~\ref{sec:perf-results}, is made available online\footnote{Sharing elimination example - application to a hyperelastic model: \url{https://gist.github.com/FabioLuporini/14e79457d6b15823c1cd}}.
\begin{figure}[h]\begin{CenteredBox}
{\subfigcapskip = 7pt \subfigure[Normal form]{\label{code:poisson_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/poisson_pre_se.code}}}
~~~~~~~~~~
{\subfigcapskip = 7pt \subfigure[After sharing elimination]{\label{code:poisson_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/poisson_after_se.code}}}
\end{CenteredBox}\caption{Applying sharing elimination to the bilinear form $J^{-T} \nabla u \cdot J^{-T} \nabla v$ arising from a Poisson equation in 2D.}\label{code:poisson}\end{figure}

>>>>

\subsubsection{Structured tensor operations}
\label{sec:se-rln}
Finite element expressions can be seen as composition of operations between tensors. We observe that, when multiplying two tensors, it often happens that the optimal scheduling strategy is to be searched in a space of size 2. To illustrate the concept, we take a matrix-vector product $y_{\Psi \Delta} = A_{\Psi} x_{\Delta}$, in which the entries of $y, A$ and $x$ depend on $\Psi, \Delta \subset \Lambda$. Let us assume that $y_{\Psi \Delta}$ itself is an operand of a larger scalar-valued expression, so its entries can represent any possible sub-expression in $\lbrace \alpha_{eij}, \beta_{eik}, \sigma_{ei} \rbrace$. For example, suppose that $A$ is the inverse Jacobian matrix of a given coefficient and $x$ the gradient of a function in two dimensions, and consider the case $\Psi = \emptyset$, $\Delta = \lbrace L_i \rbrace$. Thus an entry in $y_{L_i}$ is of the form $(a f_i + b g_i) (c f_i + d g_i) \in \sigma_{ei}$. To schedule this expression, we basically have two options:
\begin{enumerate}[label=(\roman*)]
\item Applying generalized code motion -- not needed in the example as we have one loop.
\item Searching for spatial sharing, expanding and factorizing the expression to promote temporal sharing, applying generalized code motion -- the expression in the example would be recast as $f_i (a + c) + g_i (b + d)$, with $(a + c)$ and $(b + d)$ being two hoistable sub-expressions.
\end{enumerate}
The reader can verify that (ii) improves the operation count over (i) for any size of $L_i$. In general, however, the optimal option depends on multiple factors: the loop size, the expansion cost, and code motion. One can appreciate the dichotomy between (i) and (ii) in several expressions originating from a wide range of variational forms\footnote{In fact, it systematically appears in all of the forms used for experimentation (Section~\ref{sec:perf-results}), and many more.}. We then regard as \textit{structured tensor operation} any operation between two tensors that result in sub-expressions for which there is no ambiguity in spatial sharing elimination -- implying that the optimal operation scheduling is to be determined out of two alternatives.

@@

The algorithm makes use of the following notation and terminology:
\begin{itemize}
\item $|.|$: a generic ``sizeof'' operator (e.g., cardinality of a set, extent of a loop).
\item \textit{multilinear operand}: any $\alpha_{eij}$ or $\beta_{eik}$ in the input expression.
\item \textit{multilinear symbol}: a symbol appearing within a multilinear operand depending on $L_j$ or $L_k$ (e.g., test functions, first order derivatives of test functions, etc.).
\item \textit{strategy (i)} and \textit{strategy (ii)}: the two approaches described in Section~\ref{sec:se-rln} for handling structured tensor operations.
\end{itemize}

\begin{Algo}[Sharing elimination]
\label{algo:sharing-elimination}
\normalfont
The algorithm has three main phases: initialization (step 1); scheduling of multilinear operands for reaching quasi-optimality (steps 2-4); scheduling of all other sub-expressions (step 5).
\begin{enumerate}
\item Perform a depth-first visit of the expression tree with recursive collection of identical sub-expressions.

\textit{Note: for example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$}.
\item Perform a depth-first visit of the expression tree to collect and partition multilinear operands into disjoint sets, $\mathbb{P} = \lbrace P^1, ..., P^{p}\rbrace$. $\mathbb{P}$ is such that multilinear operands in $P$ share a set of multilinear symbols $S_P$, whereas there is no sharing across different partitions.
\item For each $P \in \mathbb{P}$ such that $|P| \leq |S_P|$, apply strategy (i) to the multilinear operands.

\textit{Note: $|P|$ and $|S_P|$ represent the number of products in the innermost loop induced by $P$ if, respectively, strategies (i) or (ii) were applied. We will clarify that this directly descends from loop linearity.}

\item Build the \textit{sharing graph} $\mathbb{G} = (\mathbb{S}, \mathbb{E})$, with $S^i \in \mathbb{S}$ representing a multilinear symbol or a temporary produced at step 3. An edge $(S^i$, $S^j)$ indicates that if the sub-expressions including $S^i$ and $S_j$ were expanded, a product $S^i S^j$ would be exposed. Let $\mathbb{T}$ be the subset of vertices with a single incident edge, or ``terminals''. We apply, in order, the following rewrite rules (\textit{precondition}: \textit{action}):
\begin{enumerate}[label=(\Alph*)]
\item $\lbrace T^0, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: apply strategy (ii); $\mathbb{S} = \mathbb{S} \setminus \lbrace T^0, ..., T^n \rbrace$.\\ \textit{Note: strategy (ii) involves exposing $S (...T^0 + ... + ...T^n)$ and, subsequently, hoisting $(...T^0 + ... + ...T^n)$}. 
\item Only one terminal $T$ adjacent to $S$: apply strategy (ii) as in rule (A), but including one non-terminal symbol $S^i$. $\mathbb{S} = \mathbb{S} \setminus \lbrace T, S^i \rbrace$.
\item $\mathbb{T} = \emptyset$: take $S$, apply strategy (ii), eventually hoisting any two non-terminal symbols.\\ \textit{Note: this represents the ``fallback'' approach in presence of cyclic dependency}
\end{enumerate}
\item Perform a depth-first visit of the expression tree and, for each yet unhandled or previously hoisted sub-expression, apply the most profitable between strategies (i) and (ii).

\textit{Note: this pass speculatively assumes that structured tensor operations are in place. If the assumption does not hold, the result will generally be sub-optimal since only a subset of code motion opportunities may be exposed.}
\end{enumerate}
\end{Algo}

%TODO : dire che il costo dell'espansione è semplicemente dato da ciascun s in | S_P | times la 2 evalato al depth di quell s.
%\textit{Note: $\mathbb{S}$ is in practice very small, so this pass can be accomplished rapidly.}

\subsubsection{Examples}
\label{sec:se-examples}
Consider again Figure~\ref{code:multi_loopnest_a}. We have $\mathbb{P} = \lbrace P^0, P^1, P^2\rbrace$, with $P^0 = \lbrace b_j\rbrace$, $P^1 = \lbrace c_i\rbrace$, and $P^2 = \lbrace d_i\rbrace$. For all $P^i$, we have $|P^i| = 1 = |S_{P^i}|$, although applying strategy (i) in step 3 has no effect. The sharing graph is $\mathbb{G} = (\lbrace b_j, c_i, d_i \rbrace, \lbrace (b_j, c_i), (b_j, d_i) \rbrace$, and $\mathbb{T} = \lbrace c_i, d_i \rbrace$. Rewrite rule (4A) is hit, which leads to the code in Figure~\ref{code:multi_loopnest_c}.

Figure~\ref{code:poisson} applies Algorithm~\ref{algo:sharing-elimination} to a realistic example, the bilinear form arising from a Poisson equation in 2D. We observe that $\mathbb{P} = \lbrace P^0, P^1\rbrace$, with $P^0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P^1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$. In addition, $|P^i| = |S_{P^i}| = 2$, so strategy (i) is applied to both partitions (step 3). We then have (step 4) $\mathbb{G} = (\lbrace t^0, t^1, t^2, t^3 \rbrace, \emptyset)$, although no rewrite rule is applicable since $\mathbb{E} = \emptyset$.

\begin{figure}[h]\begin{CenteredBox}
{\subfigcapskip = 7pt \subfigure[Normal form]{\label{code:poisson_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/poisson_pre_se.code}}}
~~~~~~~~~~
{\subfigcapskip = 7pt \subfigure[After sharing elimination]{\label{code:poisson_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/poisson_after_se.code}}}
\end{CenteredBox}\caption{Applying sharing elimination to the bilinear form arising from a Poisson equation in 2D.}\label{code:poisson}\end{figure}

A much more complex example, using the hyperelastic model evaluated in Section~\ref{sec:perf-results}, is made available online\footnote{Sharing elimination example - application to a hyperelastic model: \url{https://gist.github.com/FabioLuporini/14e79457d6b15823c1cd}}
